---
title: "Final_Project"
author: "shawn"
date: "December 5, 2014"
output: html_document
---

Description:

##1. Read-in and Format Data
Converts multi sheet excel document to single data-frame with m/z,intensity, and sample name as variables. The multi-sheet excel document is the easiest and fastest way to export spectra data from the Flex Analysis program in the clinical lab. 
These read-in and data-formatting functions are useful because they are much faster than copying-and pasting columns into new spreadsheets, and this method is less error prone. These are good qualities, especially if you have a target plate with 96 samples. 

```{r}
#first, load all of the packages/libraries you will need to use this package
library(XLConnect)

#1. Converts multi sheet excel doc to single data-frame with m/z,intensity, 
#load XLConnect package to read in Excel files
importWorksheets <- function(file) {
  # filename: name of Excel file
  workbook <- loadWorkbook(file)
  sheet_names <- getSheets(workbook)
  names(sheet_names) <- sheet_names
  sheet_list <- lapply(sheet_names, function(.sheet){
    readWorksheet(object=workbook, .sheet)})
}

importWorksheets("11-20-2014_SEW.xls")

#This function converts all of the sheets from an excel file as a single list
#returns a list called spectra.list containing all sheets in excel file to global enironment
#returns sheet names so user can create dataframes containing peak lists of interest
Return.spectra.list<-function(file){
  spectra.list<-importWorksheets(file)
  spectra.list<-as.vector(spectra.list)
  spectra.names<<-names(spectra.list)
  return(spectra.names)
}

Return.spectra.list("11-20-2014_SEW.xls")

#create dataframes for spectra of interest
#could try to turn this into a function, 
#make.spectra.df<-function(list, sample) {
  #list<<-as.data.frame(list$(paste('sample')))
 # }

#not sure if that would be too hard with data frame naming scheme
#examples of how to do it manually w/out function
#strain 2398
s2398_1.1<-as.data.frame(spectra.list$'2398_1-1_0_A5_1')
s2398_1.2<-as.data.frame(spectra.list$'2398_1-2_0_B5_1')
s2398_3.1<-as.data.frame(spectra.list$'2398_3-1_0_E5_1')
#strain 630g

#strain 712
#strain F200
#strain R20291
#strain VPI

#function to format data so that dataframe contains only m/z time and intensity
#takes dataframe and sample name (can be different than dataframe name)
#as argument and deletes unnecessary info

format.spectra.df<-function(data.frame, sample, ribotype){
  data.frame<-data.frame[ ,-c(4:7,9:11)] #remove columns that aren't needed
  data.frame<-data.frame[-c(1:2), ] #remove rows that aren't needed
  names(data.frame)<-NULL #delete the variable names since they are wrong
  colnames(data.frame)<-c("m/z","time","intensity","rel.intensity") #rename
  data.frame$sample<-sample #make put the sample name in the sample column
  data.frame$ribotype<-ribotype #add the ribotype or other feature
  write.table(data.frame,file=(paste0(sample, ".txt"))) #write this to a txt file in working directory
}
####
format.spectra.df(s2398_1.1, "2398_1.1", "027")
format.spectra.df(s2398_1.2, "2398_1.2", "027")
format.spectra.df(s2398_3.1, "2398_3.1", "027")

#now you should have a separate txt file for each sample that you want to look at

#bring samples back in 
s2398_1.1<-read.table(file="2398_1.1.txt", header=T)
s2398_1.2<-read.table(file="2398_1.2.txt", header=T)
s2398_3.1<-read.table(file="2398_3.1.txt", header=T)

#add relative intensity to dataframe
relative.intens<-function(df){
  df$relative.intensity<-df$intensity/max(df$intensity)
  df<<-df
  return(df)
}
#test
rel.intens(s2398_1.1)
s2398_1.1<-s2398_1.1[ ,-c(4:5)]
#data.frame$rel.intensity<-(data.frame$intensity/(max(data.frame$intensity)))
s2398_1.1$rel.intensity<-s2398_1.1$intensity/max(s2398_1.1$intensity)
s2398_1.2$rel.intensity<-s2398_1.2$intensity/max(s2398_1.2$intensity)
s2398_3.1$rel.intensity<-s2398_3.1$intensity/max(s2398_3.1$intensity)
```
##2.Plots intensity by m/z

```{r}

plot.spectra<-function(dataframe) {
  y<-(paste0(dataframe, "$rel.intensity"))
  x<-(paste0(dataframe, "$m.z"))   
  plot(x,y, main="Relative Intensity by M/Z", xlab="M/Z", ylab=(paste0(sample,"Relative Intensity" , xlim=c(2000,13000), ylim=c(0,1))))
}

plot.spectra(s2398_1.1)

plot(s2398_1.1$rel.intensity~s2398_1.1$m.z, type="l", main="Relative Intensity by M/Z", xlab="M/Z", ylab="Relative Intensity" , xlim=c(2000,13000), ylim=c(0,1))


```
##3.Statistics to compare spectra
Does stats to compare peaks between spectra
-not sure what best measure(s) is yet…
```{r}

```
##4.Unique Peak Finding Analysis
```{r}

```
##5. Bilds Tree 

```{r}
#going to need to convert to wide format to have each value of variable (m/z) as a column, and samples as a row.
s2398_1.1$count<-c(1:51)
test<-subset(s2398_1.1, select=c("m.z","count"))
fix.shape<-reshape(test, timevar="count", idvar=c("m.z"), direction="wide")

#then the tree will make sense
#test tree
#first combine dataframes
three.spectra<-rbind(s2398_1.1$, s2398_1.2,s2398_3.1)
#put them into wide format
three.spectra<-
# prepare hierarchical cluster
hc = hclust(dist(three.spectra$m.z), "ave")
# very simple dendrogram
plot(hc)
hc = hclust(dist(s2398_1.1$m.z), "ave")
plot(hc)
```

