---
title: "Final_Project"
author: "shawn"
date: "December 5, 2014"
output: html_document
---

Description:

##1. Read-in and Format Data
Converts multi sheet excel doc to single data-frame with m/z,intensity, and sample name as variables (this is an output of the clinical lab program)
  -useful because this is the easiest way to output data from clinical lab program  	but very tedious and error prone to copy and paste all of the columns into a new 	sheet, especially if you have 96 samples.


```{r}
#1. Converts multi sheet excel doc to single data-frame with m/z,intensity, 
#load XLConnect package to read in Excel files
library(XLConnect)

importWorksheets <- function(file) {
  # filename: name of Excel file
  workbook <- loadWorkbook(file)
  sheet_names <- getSheets(workbook)
  names(sheet_names) <- sheet_names
  sheet_list <- lapply(sheet_names, function(.sheet){
    readWorksheet(object=workbook, .sheet)})
}

#This function converts all of the sheets from an excel file as a single list
#returns a list called spectra.list containing all sheets in excel file to global enironment
#returns sheet names so user can create dataframes containing peak lists of interest
Return.spectra.list<-function(file){
  spectra.list<-importWorksheets(file)
  spectra.list<<-as.vector(spectra.list)
  spectra.names<<-names(spectra.list)
  return(spectra.names)
}

#create a dataframe for spectra of interest
#could try to turn this into a function, 
#not sure if that would be too hard with data frame naming scheme
#strain 2398
#s2398_1.1<-as.data.frame(spectra.list$'2398_1-1_0_A5_1')
#s2398_1.2<-as.data.frame(spectra.list$'2398_1-2_0_B5_1')
#s2398_3.1<-as.data.frame(spectra.list$'2398_3-1_0_E5_1')
#strain 630g

#strain 712
#strain F200
#strain R20291
#strain VPI

#function to format data so that dataframe contains only m/z time and intensity
#creates a new variable called sample that can be used for merging multiple spectra
#creates new variable called ribotype where another feature can be added to dataframe
#takes dataframe and sample name (can be different than dataframe name)
#as argument and deletes unnecessary info

format.spectra.df<-function(data.frame, sample, ribotype){
  data.frame<-data.frame[-c(1:2), ]
  data.frame<-data.frame[ ,-c(4:9) ]
  names(data.frame)<-NULL
  colnames(data.frame)<-c("m/z","time",intensity","sample", "ribotype")
  data.frame$sample<-sample
  data.frame$ribotype<-ribotype
  sample<-data.frame
  write.table(data.frame, file=""sample.txt")
}
#problem! want to change name of file!!
```
##2.Plots intensity by m/z

```{r, echo=FALSE}
plot(cars)
```
##3.Statistics to compare spectra
Does stats to compare peaks between spectra
-not sure what best measure(s) is yet…
```{r}

```
##4.Unique Peak Finding Analysis
```{r}

```
##5. Bilds Tree 
based on spectra similarities 
